---
title: "CDC Director Update"
author: "ITF-SAVI"
date: "`r format(Sys.Date(), '%m/%d/%y')`"
output:
  word_document: default

---

```{r Setup, include=FALSE}
# Updates happen on Monday and Wednesday - 3-4PM
knitr::opts_chunk$set(cache=FALSE)

#detach("package:ITFAnalytics", unload = TRUE)
#devtools::install_github("randyyee/ITFAnalytics")
library(ITFAnalytics)
library(tidyverse)
library(lubridate)
library(scales)
library(sf)
library(gridExtra)
library(openxlsx)
```

```{r Data Import, include=FALSE, cache=FALSE}

c_list   <- c("Ethiopia", "Kenya", "Senegal", "Jamaica", "India")

df       <- get_covid_df()
df2      <- onetable %>% 
  right_join(df %>% select(-who_region), by = c("iso2code" = "country_code")) %>% 
  calc_add_risk()
variants <- read.xlsx(paste0("C:/Users/", Sys.getenv("USERNAME"), "/CDC/ITF-COVID19-SAVI - Documents/Routine Reports/3. Variant Maps/Inputs/20210804_Variant_Tracker_shared_with_Partners.xlsx"), sheet = 1) 
df_vax   <- get_vax()
df_man   <- get_vax_manufacturers()
```


## Global Update

- Last 7 days:
```{r Global Update Bullet #1 (Calendar Date), include=FALSE, cache=FALSE}
analysis_date   <- floor_date(Sys.Date(), "week", week_start = 7)
who_date        <- max(filter(df, source == "WHO")$date)
confirmed_cases <- sum(filter(df, source == "WHO")$new_cases, na.rm = T) 
# QC: WHO Dashboard Top
cases_pastweek  <- sum(filter(df, source == "WHO" & (date >= who_date-6))$new_cases, na.rm = T)
# QC: WHO Dashboard Table
total_deaths    <- sum(filter(df, source == "WHO")$new_deaths, na.rm = T)

top_5_cases <- filter(df, (date >= who_date - 6) & (date <= who_date)) %>%
  filter((source == "WHO" & country != "China") | source == "JHU") %>%
  group_by(country) %>%
  summarise(past_week_total = sum(new_cases, na.rm = T)/7) %>%
  ungroup() %>%
  mutate(label = paste0("(", comma(round(past_week_total)),")")) %>%
  unite(country, country, label, sep = " ") %>%
  arrange(desc(past_week_total)) %>%
  top_n(5)

last_complete_epi <- floor_date(who_date, "week", week_start = 1)-7
```
  - Global confirmed COVID-19 cases as of `r format(who_date, "%B %d")` totaled `r comma(confirmed_cases)`, with `r comma(cases_pastweek)` cases in the last 7 days. The cumulative number of deaths is `r comma(total_deaths)`.
  - Countries reporting the highest cases counts in the last week are `r top_5_cases$country[1]`, `r top_5_cases$country[2]`, `r top_5_cases$country[3]`, `r top_5_cases$country[4]`, and `r top_5_cases$country[5]`.


- Comparing week ending on `r format(analysis_date, "%B %d")` compared with the previous complete epi week:
```{r Global Update Bullet #2 (Epi Week), include=FALSE, cache=FALSE}
# QC: WHO Dashboard Top
cases_pastweek  <- sum(filter(df, source == "WHO" & (date >= floor_date(who_date, "week", week_start = 1)-7) & date < floor_date(who_date, "week", week_start = 1))$new_cases, na.rm = T)
# QC: WHO Dashboard Top
cases_prevweek  <- sum(filter(df, source == "WHO" & (date >= floor_date(who_date, "week", week_start = 1)-14 & date < floor_date(who_date, "week", week_start = 1)-7))$new_cases, na.rm = T) 
# QC: CQC
change          <- (cases_pastweek - cases_prevweek)/cases_prevweek
# QC: WHO Dashboard Top
deaths_pastweek <- sum(filter(df, source == "WHO" & (date >= floor_date(who_date, "week", week_start = 1)-7) & date < floor_date(who_date, "week", week_start = 1))$new_deaths, na.rm = T)
# QC: WHO Dashboard Top
deaths_prevweek <- sum(filter(df, source == "WHO" & (date >= floor_date(who_date, "week", week_start = 1)-14 & date < floor_date(who_date, "week", week_start = 1)-7))$new_deaths, na.rm = T) 
# QC: CQC
d_change        <- (deaths_pastweek - deaths_prevweek)/deaths_prevweek

highest_perchange <- df2 %>% 
  filter(who_region != "Other") %>%
  filter(weekdate < max(weekdate)) %>% #
  filter(weekdate == max(weekdate) | (weekdate == max(weekdate) - 7)) %>%
  group_by(region, weekdate) %>%
  summarise(Total = sum(new_cases, na.rm = T)) %>%
  ungroup() %>%
  pivot_wider(names_from = weekdate, values_from = Total) %>%
  mutate(change  = (.[[3]] - .[[2]])/.[[2]],
         label   = paste0("(", percent(change, accuracy = 0.1),")")) %>%
  unite(label, region, label, sep = " ") %>%
  arrange(desc(change))

increase_regions <- filter(highest_perchange, change > 0) %>% 
  select(label) %>% 
  lapply(., toString)
decrease_regions <- filter(highest_perchange, change < 0) %>% 
  select(label) %>% 
  lapply(., toString)

highest_perchange <- df2 %>%
  left_join(country_populations) %>%
  filter(`2020` > 10000000) %>%
  filter(weekdate < max(weekdate)) %>% #
  filter(weekdate >= max(weekdate) - 7) %>%
  group_by(country, weekdate) %>%
  summarise(Total = sum(new_cases, na.rm = T)) %>%
  ungroup() %>%
  pivot_wider(names_from = weekdate, values_from = Total) %>%
  mutate(change = (.[[3]] - .[[2]])/.[[2]],
         label   = paste0("(", percent(change, accuracy = 0.1),")")) %>%
  unite(label, country, label, sep = " ") %>%
  arrange(change)

increase_countries <- highest_perchange %>%
  filter(change > 0 & !is.infinite(change)) %>% 
  top_n(n = 5, wt = change) %>% 
  select(label) %>% 
  lapply(., toString)
```
  - Overall cases `r ifelse(change > 0, paste("increased", percent(change, accuracy=0.1)), paste("decreased",percent(change, accuracy=0.1), "decrease"))` and deaths `r ifelse(d_change > 0, paste("increased", percent(d_change, accuracy=0.1)), paste("decreased",percent(d_change, accuracy=0.1), "decrease"))`.
  - By region, cases increased in the following regions: `r increase_regions`. Cases decreased in the following: `r decrease_regions`.
  - Countries with highest percent increase in cases were `r increase_countries`.


- Vaccination
```{r Vaccination Bullets (Epi Week), include = FALSE, cache=FALSE}

df_vax <- df_vax %>%
  mutate(people_vaccinated_per_hundred_cat = cut(people_vaccinated_per_hundred, breaks = c(0, 1, 3, 10, 30, Inf)),
         people_fully_vaccinated_per_hundred_cat = cut(people_fully_vaccinated_per_hundred, breaks = c(0, 1, 3, 10, 30, Inf))) %>%
  full_join(country_coords, by = c("iso_code" = "iso3code")) %>% 
  group_by(location) %>% 
  filter(!is.na(people_vaccinated_per_hundred)) %>%
  arrange(date) %>%
  top_n(1, date) %>%
  distinct(iso_code, .keep_all = TRUE)

vax_date      <- max(df_vax$date)
vax_countries <- length(unique(df_vax$location))
vax_total     <- sum(filter(df_vax, date == max(date))$total_vaccinations, na.rm = T)
```
  - Vaccination is reported in at least `r vax_countries` countries or territories. As of `r format(vax_date, "%B %d")`, a total of `r comma(vax_total)` doses have been administered.


- Global Epidemiology
```{r Epi Curve (Epi Week), echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE, fig.height = 7, fig.width = 11}
#QC: WHO Dashboard
plot_epicurve(filter(df, source == "WHO" & region != "Other" & date <= analysis_date))
```


- Variants
```{r VOC Bullet, include = FALSE, cache=FALSE}
# QC Tracker Pivots
variants <- variants %>%
  full_join(country_coords, by = c("ISO" = "iso3code")) %>%
  filter(!is.na('WHO assiged name')) %>%
  st_as_sf()

max_variant_date <- analysis_date
n_alpha <- filter(variants, `WHO.assigned.name` == "Alpha") %>% nrow()
n_beta  <- filter(variants, `WHO.assigned.name` == "Beta")  %>% nrow()
n_gamma <- filter(variants, `WHO.assigned.name` == "Gamma") %>% nrow()
n_delta <- filter(variants, `WHO.assigned.name` == "Delta") %>% nrow()
```
  - Based on data from `r format(max_variant_date, "%B %d")`, Delta (B.1.617.2) has been identified in `r n_delta` countries and is present in all regions.

## Countries we are watching (table with detailed data follows the country summaries)
```{r Epi Curve Individual, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE, fig.height = 3, fig.width = 18}
c_list <- c("Japan")
for (c in c_list){

  grid.arrange(
    plot_epicurve_ind(filter(df2, country == c) %>% rename(cases  = week_case_incidence),
                      type = "cases",  incidence = T),
    plot_epicurve_ind(filter(df2, country == c) %>% rename(deaths = week_death_incidence),
                      type = "deaths", incidence = T),
    ncol = 2
  )

}

```

```{r country table, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3, fig.width = 15}

table_countriesofconcern(df2, df_vax, df_man, c_list)
```

## MORE DETAILED DATA ON VACCINATION AND VARIANTS: 

## Vaccination

Vaccination is reported in at least `r vax_countries` countries or territories. As of `r format(vax_date, "%B %d")`, a total of `r comma(vax_total)` doses have been administered.

```{r vaccination map, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE, fig.height = 7, fig.width = 11}
  cat_vals = c("#d4ece8","#a2d9d2", "#1f9fa9", "#005e70", "#27343a")
  cat_labs = c("<1", "1- <3", "3- <10", "10- <30", "30+")

  map_template(rename(df_vax, result = people_vaccinated_per_hundred_cat), cat_labs, cat_vals) +
    labs(title    = paste0("People Vaccinated per 100 People, ", format(max(df$date), "%B %d, %Y")),
         subtitle = "Number of people out of 100 who received at least one vaccine dose; does not represent percent of \npopulation fully vaccinated",
         caption  = "Note:
       -Countries in white do not have data reported for total people vaccinated
       -Vaccine data are incomplete and data may be out of date")+ 
  guides(fill = guide_legend(title="People Vaccinated per 100"))


    cat_vals = c("#CCECE6","#99D8C9", "#66C2A4", "#2CA25F", "#006D2C")
    cat_labs = c("<1", "1- <3", "3- <10", "10- <30", "30+")

  map_template(rename(df_vax, result = people_fully_vaccinated_per_hundred_cat), cat_labs, cat_vals) +
      labs(title    = paste0("People Fully Vaccinated per 100 People, ", format(max(df$date), "%B %d, %Y")),
           subtitle = "Represents percent of population fully vaccinated",
           caption  = "Note:
       -Countries in white do not have data reported for fully vaccinated
       -Vaccine data are incomplete and data may be out of date") +
  guides(fill = guide_legend(title="People Fully Vaccinated per 100"))
```

## Variants of Concern

- Alpha(B.1.1.7) has been identified in `r n_alpha` countries.
- Beta(B.1.351) has been identified in `r n_beta` countries.
- Gamma(P.1) has been identified in `r n_gamma` countries.
- Delta(B.1.617.2) has been identified in `r n_delta` countries.

```{r Variant Maps, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE, fig.height = 7, fig.width = 11}
variants$Verification <- ordered(variants$Verification, 
                                 levels = c("Verified", "Not verified"),
                                 labels = c("Verified by WHO", "Reported from\nother sources"))
grid.arrange(
  
  # Variant Alpha (B.1.1.7) Map
  map_template(filter(variants, `WHO.assigned.name` == 'Alpha') %>% 
                 mutate(result = Verification),
               #country_coords, 
               c("Verified by WHO", "Reported from\nother sources"), 
               c("#2171b5","#94939e", "grey95")) +
    labs(title = "Countries reporting Alpha (B.1.1.7) variant", fill = ""),
  
  # Variant Beta (B.1.351) Map
  map_template(filter(variants, `WHO.assigned.name` == 'Beta') %>% 
                 mutate(result = Verification), 
               #country_coords, 
               c("Verified by WHO", "Reported from\nother sources"),
               c( "#238b45","#94939e", "grey95")) +
    labs(title = "Countries reporting Beta (B.1.351) variant", fill = ""),
  
  # Variant Gamma (P.1) Map
  map_template(filter(variants, `WHO.assigned.name` == 'Gamma') %>% 
                 mutate(result = Verification), 
               #country_coords, 
               c("Verified by WHO", "Reported from\nother sources"), 
               c("#6a51a3","#94939e", "grey95")) +
    labs(title = "Countries reporting Gamma (P.1) variant", fill = ""), 
  
  # Variant Delta (B.1.617.2) Map
  map_template(filter(variants, `WHO.assigned.name` == 'Delta') %>% 
                 mutate(result = Verification), 
               # country_coords, 
               c("Verified by WHO", "Reported from\nother sources"),
               c( "#e30b5d","#94939e", "grey95")) +
    labs(title = "Countries reporting Delta (B.1.617.2) variant", fill = ""),
  
  ncol = 2
)
```



